---
title: "Practical Day 4"
author: "Natalie Chidikwindi"
format: html
editor: visual
---

```{r}
# Install tidyverse if not already installed
if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}
library(tidyverse)

# Install and load nycflights13 for flight data
if (!requireNamespace("nycflights13", quietly = TRUE)) {
  install.packages("nycflights13")
}
library(nycflights13)
# installed the tidverse packageDescription
```


## Question 1

Display the `flights` dataset in an alternative format to simply printing it (i.e. running `flights`).

### Answer:

```{r}

data("flights")
flights
```

## Question 2

Rewrite the code using `dplyr` and the pipe:

### Answer:

```{r}

carrier_dist_stats <- flights %>% 
  filter(month == 1) %>% #filter for the first month
  group_by(carrier) %>% # ensures output shows by carrier the mean and sd distance in average
  summarise(
    mean_distance = mean(distance, na.rm = TRUE),
    sd_distance   = sd(distance, na.rm = TRUE),
    
  ) %>% 
  arrange(mean_distance, sd_distance)

carrier_dist_stats

```

## Question 3

Explain why the standard deviation is `NA` for one carrier, and why it is `0` for others. Demonstrate your answer using code.

### Answer:

```{r}
data <- flights # loaded to variable called data
```

```{r}
# looks at the mean of the yv carrier and compares the number or rows in the vector with the number of rows with duplicated values for distance
yv_dist_vec <- flights %>% 
  filter(month ==1) %>% 
  filter(carrier == "YV") %>% 
  select(distance)
rownum <- as.numeric(length(t(yv_dist_vec)))
rownum
duplicated_rows <- sum(duplicated(yv_dist_vec)) 
duplicated_rows+1 # added 1 because it compares with the first row, discounting it as a duplicate
```

all the distance values for the carrier YV are equal to the mean distance for that carrier. Therefore there is no variance and therefore the standard deviation is 0.

```{r}
#looks at the number of records of distance for OO carrier
oo_dist_vec <- flights %>% 
  filter(month ==1) %>% 
  filter(carrier == "OO") %>% 
  select(distance)
rownum <- as.numeric(length(t(oo_dist_vec)))
rownum
```

There is only 1 recorded distance value for the carrier OO, therefore there variance and standard deviation are not applicable

## Question 4

Using `tidyr` and `dplyr` where appropriate, construct a dataframe where the carriers are along the columns, and the rows are the average departure delay (`dep_delay`) flown by each carrier (`carrier`) in each month.

### Answer:

```{r}
# creates a tibble with average depature delay for each carrier
avg_delay_table <- flights %>% 
  group_by(carrier, month) %>% 
  summarise(avg_dep_delay = mean(dep_delay, na.rm = TRUE)) %>% 
  pivot_wider(
    names_from = carrier,
    values_from = avg_dep_delay
  )

avg_delay_table
```

## Question 5

Calculate the proportion of flights that were delayed (`dep_delay` greater than 0) but arrived on or before time (`arr_delay` less than or equal to 0).

### Answer:

```{r}
#creates a tibble with flights satisfying the condition
delay_table <- flights %>% 
  filter(dep_delay > 0 & arr_delay <=0)
# calculates the proportion as a percentage
delayed_num <-nrow(delay_table)
delayed_prop <- (delayed_num / nrow(flights)) *100

round(delayed_prop)

```

11% of flights were delayed but arrived on/before time.

## Question 6

Using the `airlines` and `flights` datasets, do the following, showing the output from each step:

Identify routes that more than one airline flies

### Answer:

```{r}

#merged the two tables 
airline_flight <- left_join(flights,airlines, by = "carrier")

route <- airline_flight %>% 
  group_by(origin, dest) %>% 
  summarise(carriers_num = n_distinct(carrier)) %>% 
  filter(carriers_num > 1) # filters to show only routes that have more than 1 carrier
route


  

```

For each such route, identify the airline with the worst and best average arrival delay.

### Answer:

```{r}
avg_arr_delay <- airline_flight %>% 
  group_by(origin, dest, carrier, name) %>% 
  summarise(mean_delay = mean(arr_delay, na.rm = TRUE)) 
avg_arr_delay

```

For each such route, identify the airline with the worst and best average arrival delay.

### Answer:

```{r}
best <- avg_arr_delay %>% 
  group_by(origin,dest,name) %>% 
  summarise(best_arr = min(mean_delay))

worst <- avg_arr_delay %>% 
  group_by(origin,dest,name) %>% 
  summarise(worst_arr = max(mean_delay))


best_worst <- left_join(best,worst, by = c("origin" = "origin", "dest" = "dest"))
best_worst
  
```

Identify the route with the greatest difference between the best and worst performing airlines

### Answer:

```{r}
greatest_diff <- best_worst %>% 
  group_by(origin,dest) %>% 
  mutate(difference = worst_arr - best_arr) %>% 
  arrange(desc(difference))
head(greatest_diff, n = 1)
```

Determine the reason for this difference

### Answer:

```{r}
airline_flight %>% 
  filter(origin == "JFK", dest == "ATL") %>% 
  group_by(carrier,name) %>% 
  summarise(num_flights = n(), avg_dep_delay = mean(dep_delay, na.rm = TRUE), mean_arr_delay = mean(arr_delay)) %>% 
  relocate(carrier, name)
  
```

The EV carrier only has 1 flight that was delayed by 124 minutes, therefore the average delay will be higher than the other carriers that have had multiple flights.

## Question 7

Identify all columns with missing entries, typos and any other inconsistencies in the dataset below

### Answer:

```{r}
df <- structure(list(id = c("id_1", "id_2", "id_3", "id_4", "id_5", 
"id_6", "id_7", "id_8", "id_9", "id_10", "id_11", "id_12", "id_13", 
"id_14", "id_15", "id_16", "id_17", "id_18", "id_19", "id_20", 
"id_21", "id_22", "id_23", "id_24", "id_25", "id_26", "id_27", 
"id_28", "id_29", "id_30", "id_31", "id_32", "id_33", "id_34", 
"id_35", "id_36", "id_37", "id_38", "id_39", "id_40", "id_41", 
"id_42", "id_43", "id_44", "id_45", "id_46", "id_47", "id_48", 
"id_49", "id_50"), age = c(50L, 34L, 70L, 33L, 22L, 61L, 69L, 
73L, 62L, 56L, 71L, 33L, 73L, 44L, 45L, 46L, 24L, 70L, 46L, 76L, 
47L, 76L, 28L, 48L, 54L, 27L, 45L, 26L, 61L, 28L, 38L, 55L, 33L, 
36L, 62L, 58L, 72L, 31L, 34L, 51L, 61L, 64L, 26L, 28L, 60L, 29L, 
42L, 46L, 79L, 72L), gender = c("male", "male", "male", "female", 
"female", "male", "female", "male", "male", "female", "female", 
"male", "male", "female", "male", "male", "male", "male", "female", 
"male", "male", "male", "male", "female", "femal", "male", "female", 
"female", "female", "female", "male", "female", "female", "female", 
"male", "male", "female", "male", "female", "female", "male", 
"female", "female", "male", "male", "female", "male", "male", 
"male", "female"), height = c(174.4, 197.7, 174.1, 194.5, NA, 
180.4, 170.5, 157.4, 196.8, 165.1, 153, 197.4, 186, 157.1, 177.5, 
197.7, 179.3, 170.2, 182.4, NA, 165.4, 161, 168.5, 199.2, 157.7, 
154.6, 157.1, 184.5, 181, 194.6, 183.6, 186.9, 176.1, 183, 191.1, 
189.3, 199, 172, 165.6, 170.5, 150.5, 159.2, 192.1, 161.6, 162, 
153.8, 162.3, 186.6, 192.4, 174.9), weight = c(69.4, 62.3, 55.6, 
69.5, 78.6, 60.8, 72.2, 60.9, 75.1, 67.7, 82.5, 68.7, 67.8, 76.7, 
87, 61.1, 70.6, 63.3, 81.5, 59.2, 93.2, 87.3, 83.4, 80.9, 68.6, 
76.5, 93.7, 79.1, 92, 65.6, 85.4, 63.3, 79.7, 74.1, 63.3, 78.2, 
95.7, 95.1, 63.7, 66.1, 99.3, 81, 96.9, 73.3, 70.3, 83, 57.6, 
78.6, 61.9, 98.1), blood_type = c("O", "A", "O", "O", "B", "AB", 
"O", "O", "O", "AB", "A", "O", "O", "O", "B", "A", "B", "AB", 
"O", "AB", "A", "AB", "O", "B", "A", "A", "B", "AB", "A", "B", 
"B", "A", "O", "O", "O", "B", "O", "A", "A", "B", "A", "O", "AB", 
"A", "A", "O", "O", "B", "A", "O"), disease_status = c("diseased", 
"healthy", "healthy", "healthy", "healthy", "healthy", "diseased", 
"healthy", "diseased", "Healthy", "diseased", "healthy", "diseased", 
"healthy", "diseased", "healthy", "healthy", "healthy", "healthy", 
"healthy", "healthy", "diseased", "healthy", "diseased", "healthy", 
"healthy", "healthy", "healthy", "diseased", "diseased", "healthy", 
"healthy", "healthy", "diseased", "diseased", "diseased", "healthy", 
"diseased", "healthy", "healthy", "healthy", "healthy", "healthy", 
"diseased", "diseased", "diseased", "healthy", "healthy", "diseased", 
"diseased"), cholesterol = c(228, 223, 213, 198, 166, 151, 195, 
199, 189, 196, 221, 156, 185, 230, 234, 174, 185, 236, 235, 180, 
165, 220, 160, 153, 250, 153, 184, 242, 212, 179, 224, 233, 181, 
199, 220, 214, 214, 248, 191, 162, 203, 173, 199, 187, 248, 189, 
173, 212, 164, 247), glucose = c(96, 78, 101, 119, 103, 91, 86, 
NA, 77, 80, 115, 85, 88, 109, NA, 71, 90, 94, 91, 87, 113, 93, 
97, 118, 109, 80, 85, 119, 99, 108, 89, 108, 97, 116, 79, 84, 
75, 81, 119, NA, 106, 109, 75, 82, 84, 75, 76, 120, 119, 77), 
    smoker = c("yes", "yes", "yes", "yes", "no", "yes", "no", 
    "yes", "no", "no", "no", "no", "no", "yes", "no", "yes", 
    "yes", "yes", "yes", "yes", "yes", "yes", "yes", "yes", "no", 
    "no", "yes", "yes", "yes", "no", "no", "yes", "no", "yes", 
    "no", "yes", "no", "yes", "yes", "yes", "no", "no", "yes", 
    "no", "no", "no", "no", "no", "no", "yes"), exercise = c("occasional", 
    "regular", "occasional", "regular", "none", "occasional", 
    "regular", "none", "occasional", "none", "occasional", "none", 
    "none", "regular", "occasional", "none", "regular", "regular", 
    "none", "occasional", "none", "occasional", "occasional", 
    "occasional", "regular", "occasional", "regular", "regular", 
    "regular", "occasional", "occasional", "none", "none", "regular", 
    "occasional", "occasional", "none", "none", "none", "none", 
    "occasional", "regular", "regular", "none", "regular", "occasional", 
    "occasional", "none", "occasional", "regular"), income = c(84820L, 
    81547L, 22588L, 72490L, 74533L, 25338L, 41469L, 57315L, 63629L, 
    88662L, 62615L, 56261L, 58499L, 82232L, 77584L, 77275L, 38468L, 
    54510L, 91326L, 78611L, 31402L, 29586L, 21441L, 58269L, 84173L, 
    88295L, 37940L, 43750L, 69750L, 92356L, 82518L, 91455L, 68866L, 
    51178L, 68275L, 27689L, 35418L, 81318L, 62405L, 86851L, 25654L, 
    47553L, 74474L, 51409L, 22607L, 55360L, 96351L, 21516L, 41927L, 
    55810L), education = c("master", "bachelor", "PhD", "master", 
    "bachelor", "highschool", "PhD", "highschool", "PhD", "PhD", 
    "bachelor", "highschool", "master", "bachelor", "PhD", "PhD", 
    "PhD", "bachelor", "master", "highschool", "PhD", "highschool", 
    "bachelor", "master", "highschool", "highschool", "master", 
    "master", "bachelor", "PhD", "highschool", "PhD", "master", 
    "master", "master", "PhD", "highschool", "master", "master", 
    "highschool", "bachelor", "highschool", "bachelor", "PhD", 
    "bachelor", "highschool", "master", "highschool", "bachelor", 
    "bachelor"), region = c("North", "South", "North", "West", 
    "North", "West", "South", "South", "West", "South", "West", 
    "South", "West", "East", "North", "West", "North", "North", 
    "West", "North", "East", "West", "South", "North", "North", 
    "East", "East", "North", "North", "West", "South", "West", 
    "West", "East", "West", "North", "West", "North", "East", 
    "North", "West", "South", "South", "East", "North", "West", 
    "West", "East", "North", "East"), marital_status = c("divorced", 
    "single", "divorced", "divorced", "divorced", "divorced", 
    "divorced", "married", "divorced", "married", "divorced", 
    "widowed", "married", "single", "widowed", "widowed", "single", 
    "divorced", "widowed", "widowed", "single", "married", "single", 
    "married", "widowed", "married", "single", "single", "widowed", 
    "married", "widowed", "divorced", "single", "married", "single", 
    "widowed", "widowed", "married", "widowed", "divorced", "married", 
    "married", "divorced", "single", "married", "widowed", "divorced", 
    "divorced", "single", "divorced")), row.names = c(NA, -50L
), class = c("tbl_df", "tbl", "data.frame"))
```

```{r}
colSums(is.na(df)) # looked for na values in the dataset 
unique(df$gender) # looked for unique ways gender was writtern
unique(df$disease_status) # did the same thing with disease status as for gender
sapply(df, function(x) if(is.character(x)) unique(x))
# The created a tibbles to show all the rows this inconsistencies
issues <- df %>%
  filter(
    is.na(height) |
    is.na(glucose) |
    gender == "femal" |
    disease_status == "Healthy"
  )

issues
```
