---
title: "Day 3 Practical"
author: "Natalie Chidikwindi"
format: html
editor: visual
---

## Practical 3

These are my answers for the last 2 questions

## Question 4

Plot the function $f(x) = \sin(x)$ for $x \in [-2, 2]$.

### Answer:

```{r}
x <- seq(-2,2, 0.01)
sin_fx <- sin(x)
plot(x, sin_fx, "l",
     main ="plot of sin_f(x) function")
```

## Question 5

Generate 1000 values from a $t$-distribution with 1 degree of freedom.

Construct a QQ-plot with 95% probability bands (envelopes) to check whether the above values come from a standard normal distribution. You can only use `rnorm()`, but not `qnorm`, `dnorm`, `pnorm`, or R's qqplot functions.

### Answer:

```{r}
set.seed(142) # to ensure the t values are reproducable
# sorted both the t values and normal values in order
t_values <- sort(rt( n = 1000, df = 1))
norm_values <- sort(rnorm(1000))

# plotted the t values and a linear line 
plot(norm_values, t_values,
     xlab = "Standard Normal Quantiles",
     ylab = "Sample Quantiles (t df=1)",
     main = "QQ-plot of t-distributed values with a 95% confidence band",
     ylim = c(-5, 5))
abline(0, 1, col = "red") 

# simulated 1000 samples of normal values with 1000 in each sample
sim_num <- 1000
sim_norm <- replicate(sim_num, sort(rnorm(1000)))
# Calculated confidence bands at 95% significance based on normal values 
lower <- apply(sim_norm, 1, quantile, probs = 0.025)
upper <- apply(sim_norm, 1, quantile, probs = 0.975)
# Plottes the lines above
lines(norm_values, lower, col = "blue", lty = 2)
lines(norm_values, upper, col = "blue", lty = 2)


```

The t-values are not within the confidence and therefore we can conclude that they are not normally distributed.
