---
title: "Day 1 Practical"
author: "Natalie Chidikwindi"
format: html
editor: visual
---

## Question 1

Find all rows in “airquality” that have missing values. Note that the `airquality` dataset in `R` is always available

### Answer:

```{r}
#loaded in dataset for airquality, it is already available in r
data <- airquality
head(data)
```

```{r}
#the function runs through all the rows and columns to find the rows which contains a NA vaue and appends the row number to the matrix null_rows

null_rows <- data[rowSums(is.na(data))>0,] 
as.integer(rownames(null_rows)) # rownames outputs a string value for the row number but I wanted it as a integer typ so i converted it
```

## Question 2

Find mean, sd, min, max for each of temperature and ozone level.

### Answer:

There are NA values in our dataset, so they need to be accounted for before we can find some summary statistics for Ozone and Temperature in our data.

```{r}
# For Ozone

cat(" Ozone statistics: \n mean = ", round(mean(data$Ozone, na.rm = TRUE),2), "\n",
"standard deviation = ", round(sd(data$Ozone, na.rm = TRUE),2), "\n", 
"min = " , min(data$Wind, na.rm = TRUE), "\n",
"max = ", max(data$Ozone, na.rm = TRUE),"\n")

# For Temperature

cat("\n Temperature statistics: \n mean = " , round(mean(data$Temp, na.rm = TRUE),2), "\n",
"standrard deviation =" ,  round(sd(data$Temp, na.rm = TRUE),2), "\n",
"min = ", min(data$Temp, na.rm = TRUE), "\n",
"max = ", max(data$Temp, na.rm = TRUE))


```

## Question 3

Fit a simple linear regression model to these data, i.e. find the $beta$ estimates, using the equation above, and matrix calculations in R

### Answer:

```{r}
#loading of dataset
car_data <- cars

```

```{r}
# assignment of Y and design matrix(X)
X <- cbind(1,car_data$dist) # made the design matrix X and added a column of 1's for the beta estimate for the constant in the linear regression equation
Y <- car_data$speed
```

```{r}
# manually sovling for bhat vector
b_estimate <- solve(t(X)%*%X)%*%t(X)%*%Y
round(b_estimate, 4)
```

## Question 4

Check that you get the same estimates as when fitting the linear regression model using lm() in R.

### Answer:

```{r}
model <- lm(Y~X, data = car_data)
round(model$coefficients,4)
```

The beta estimates from the $lm()$ are the same as the ones from the manual calculation in question 3.
