---
title: "Prac day 12"
author: "Natalie Chidikwindi"
format: html
editor: visual
---

```{r}
#| echo: true
#| results: hide
#| message: false
#| warning: false
#Loading of all libraries needed for the prac
library(foreach)
library(tidyverse)
library(MASS)
library(doParallel)
library(knitr)
library(kableExtra)
```

## Question 1

Use a `foreach` loop to repeat 100 times:

generate a random sample from an exponential distribution with mean 1

calculate mean and variance

row-bind your results (rbind) (results = mean and variance).

### Answer

```{r}
#creates a column vector of all the means and variances for each 100 values sampled from an exponential distribution with mean 1 for each iteration
result =  foreach(i = 1:100, .combine = "rbind") %do%{
   exp_vec = rexp(100, 1)
   mean = mean(exp_vec)
   var = var(exp_vec)
   c(round(mean,2),round(var,2)) 
}
# changes the column and row names
colnames(result) <- c("Mean", "Variance")
rownames(result) <- c(seq(1,100))

#Put output into a table using kable
kable(result, caption = "Table 1 : Means and Variances for 100 exponential distributions with mean 1") %>% 
  kable_styling(bootstrap_options = "striped", "hover") %>% 
  scroll_box(., height = "300px")
```

## Question 2

Use the `doParallel` package and `foreach` to bootstrap the median for the galaxies data (in library MASS).

### Answer

```{r}
set.seed(123)
c1 <- makeCluster(2) # assigned 2 workes to the task 
registerDoParallel(c1)
# gives time output of how long it will take both workers to create a vector with the boostrapped values
system.time({median_estimate <- median(galaxies)
  boostrap <- foreach(i =  1:1000, .combine = 'rbind', .packages = 'MASS') %dopar%{
    cases = sample(galaxies, size = 82, replace = T)
    median = median(cases)
    c(median)
}}) 

stopCluster(c1)
colnames(boostrap) <- c("Boostrap Median")
rownames(boostrap) <- c(seq(1,1000))
boostrap %>% 
  kable() %>% 
  kable_styling(bootstrap_options = "striped", "hover", full_width = FALSE) %>% 
  scroll_box(., height = "300px")

```

Try doing 1000 bootstrap samples at a time instead of managing single bootstrap samples.

### Answer

```{r}
set.seed(123)
c1 <- makeCluster(2) #assigned 2 workers
registerDoParallel(c1)
median_estimate <- median(galaxies)
system.time({
  boostrap2 <- foreach(i =  1:10, .combine = 'cbind', .packages = 'MASS') %dopar%{
    median_short = replicate(100, median(sample(galaxies, size = 82, replace = T)))
    c(median_short)
    
}})

stopCluster(c1)

colnames(boostrap2) <- c(seq(1,10))
rownames(boostrap2) <- c(seq(1,100))
boostrap2 %>% 
  kable(caption = "Boostrap values of 1000 values for each iteration") %>% 
  kable_styling(bootstrap_options = "striped", "hover", full_width = FALSE ) %>% 
  scroll_box(., height = "300px")

```

the time doing 1000 at a time was faster than the time doing each bootstrap iteration.

## Question 3

Estimate coverage of a percentile bootstrap confidence interval for the following scenario: sample of size 50 from an exponential distribution with mean 1.

### Answer

The true mean estimate has been given which is 1

```{r}
mean_estimate <- 1
```

Coverage after bootstrapping

```{r}
set.seed(123)
c2 <- makeCluster(5) # assigned 5 workers 
registerDoParallel(c2)
# calculated coverage for each 1000 boostraps
coverage <- foreach(i = 1:1000, .combine = "rbind") %dopar%{
  values = rexp(50,1)
  cases = replicate(1000, mean(sample(values, size = 50 , replace = T)))
  p_CI = quantile(cases, probs = c(0.025, 0.975))
  c(1 >= p_CI[1] && 1 <= p_CI[2])
}

stopCluster(c2)

mean(coverage)
```

## Question 4

The package iterators provides several functions that can be used to create sequences for the `foreach` function. For example, the `irnorm` function creates an object that iterates over vectors of normally distributed random numbers. It is useful when you need to use random variables drawn from one distribution in an expression that is run in parallel.

In this exercise, use the `foreach` and `irnorm` functions to iterate over 3 vectors, each containing 5 random variables. Find the largest value in each vector, and print those largest values.

Before running the `foreach` function set the seed to 1234.

### Answer

```{r}
set.seed(1234)
# creates 3 vectors with 5 vales from a normal distribution and then 
Large <- foreach(i = 1:3, .combine = 'rbind') %do% {
  vec = irnorm(n =5, count = 3)
  value = foreach(x = vec, .combine = c) %do% {
    x                                           # extracts each value x in each iteration
  }
  largest = max(value)
  c(round(largest, 2))             # Extracts the largest value from each iteration 
}

#changed row and column names
colnames(Large) <- c("Largest Value")
rownames(Large) <- c("Vector 1", "Vector 2", "Vector 3")

#produces a table with the output
Large %>% 
  kable(caption = "Table 2: Largest values of 3 vectors with normally distributed values") %>% 
  kable_styling(bootstrap_options = "striped", "hover", full_width = T)

```

## Question 5

Compare run time between `parLapply`, `foreach` and `replicate` for the above problem.

Using `foreach`

```{r}
set.seed(1234)
#stores the time it takes to to Q4 using foreach
time1<- system.time({
  foreach(i = 1:3, .combine = c) %do% {
    vec = irnorm(n =5, count = 3)
    value = foreach(x = vec, .combine = c) %do% {
      x
    }
    largest = max(value)
    c(largest)
}})
```

using `replicate`

```{r}
set.seed(1234)
#stores the time it takes to use replicate method for Q4
time2 <- system.time({
largest = replicate(3, max(rnorm(5)))
})
```

using `parLappy`

```{r}
# stores the time it takes to use the parLapply function for Q4
set.seed(1234)
time3 <- system.time({
c3 <- makeCluster(2)
rnorm_max <- function(i){
  values = rnorm(5)
  return(max(values))
}

largest <- parLapply(c3, as.list(1:3), rnorm_max)

stopCluster(c3)
})

```

Table to compare times

```{r}
times <- rbind(time1, time2, time3)
times <- times [,-c(4,5)]
# changes column and rownames
colnames(times) <- c("user", "system", "elapsed")
rownames(times)<- c("foreach", "replicate", "parLapply")

times %>% 
  kable(caption = "Table 3: Time usage for each method") %>% 
  kable_styling(bootstrap_options = "striped", "hover") 
```

From Table 3 , the fastest method is using `replicate` and the method with the longest waiting time is `parLapply.`
